Naturalness in U(1) extensions of the MSSM

### 2015-01-20 paper/e6ssm-tuning-paper.tex
Merged some of the U(1) and E6 literature review into model section and hopefully thereby reduced repitition and improved clarity. may still be a bit waffly.

Here we will refer to U(1) extensions of the MSSM, which solve the
$\mu$-problem as the USSM \cite{Cvetic:1995rj, Jain:1995cb,
  Nir:1995bu, Cvetic:1996mf, Cvetic:1997ky}, but since avoiding gauge
anomalies is straightforward in the when the $U(1)$appears from the
breakdown of $E_6$ and complete matter multiplets are present at low
energies, we will sometimes refer to such models as $E_6$ inspired.
Models with with complete $E_6$ matter multiplets at low energies and
an extra $U(1)$ under which the right handed neutrino remains
interaction-less are often refered to as the exceptional supersymmtric
standard models (E$_6$SSM). However we will reserve the name E$_6$SSM
to the original formulation of models
\cite{King:2005jy,King:2005my,Athron:2010zz}, and refer to similar
models as E$_6$SSM-variants\cite{Howl:2007zi, Braam:2009fi,
  Braam:2010sy, Hall:2011zq, Nevzorov:2012hs, Athron:2014pua}.

In this paper we will often focus on a very well motivated $E_6$
inspired $U(1)$ extension, the Exceptional Supersymmetric Standard
Model (E$_6$SSM)\cite{King:2005jy,King:2005my,Athron:2010zz}.  In this
model since the right handed neutrino remains interaction-less they
may explain the tiny observed masses of neutrinos via the see-saw
mechanism and the baryon asymmetry in the universe via leptogenesis
\cite{Hambye:2000bn,King:2008qb, King:2008gw}.  Recently it has also
been studied in the context of electroweak baryogenesis
\cite{Chao:2014hya}.

The gauge coupling running in the E$_6$SSM at the two-loop level leads
to unification more precisely than in the MSSM \cite{King:2007uj} or
in slightly modified scenarios two-step unification can take place
\cite{Howl:2007hq,Howl:2007zi}. If the exotic particles are light in
these models this can open up non-standard decays of the SM--like
Higgs boson \cite{Hall:2010ix,Nevzorov:2013tta,Athron:2014pua}. The
correct relic abundance could be obtained entirely through an almost
decoupled ``inert'' neutralino sector \cite{Hall:2009aj},.  However
this is no longer phenomenologically viable due to limits from direct
detection of dark matter
\cite{2011PhRvL.107m1302A,2012PhRvL.109r1301A, Akerib:2013tjd} and due
to a significant suppression of the decay of the lightest Higgs boson
into SM states, due to a new channel into inert singlinos opening up.

The impact of gauge kinetic mixing in the case where both of the extra
U(1)'s appearing from the breakdown of E$_6$ are present at low energy
was studied in \cite{Rizzo:2012rf}.  The E$_6$SSM was also included in
a studies looking at how first or second generation sfermion masses
can be used to constrain the GUT scale parameters \cite{Miller:2012vn}
and the renormalization of VEVs \cite{Sperling:2013eva,
  Sperling:2013xqa}.  The particle spectrum and collider signatures of
constrained version of the E$_6$SSM have been studied in a series of
papers,
\cite{Athron:2009ue,Athron:2009bs,Athron:2011wu,Athron:2012sq}. The
threshold corrections to the $\overline{DR}$ gauge and Yukawa
couplings in the E$_6$SSM were calculated and the numerical impact in
the constrained version examined \cite{Athron:2012pw}.

@@

  The couplings for the $U(1)^\prime$ gauge group should allow the
  following renormalisable superpotential terms required in the USSM,
  \be W_{USSM} = y^U_{ij} \hat{u}^c_i \hat{H}_u \hat{Q}_j + y^D_{ij}
  \hat{d}^c_i \hat{H}_d \hat{Q}_j + y^E_{ij} \hat{e}^c_i \hat{H}_d
  \hat{L}_j + \lambda \hat{S} \hat{H}_d \hat{H}_u, \ee with $i,j \in
  \{1,2,3\}$.

The $U(1)^\prime$ charges should allow for cancellations of gauge
anomalies. The most elegant way to do this is to use an extra $U(1)$
gauge symmetry that can be obtained from the break down of the $E_6$
gauge symmetry which is anomaly free and have all matter multiplets
from the $Z7$-plet representation of $E_6$ survive down to low energies.

@@

To ensure anomaly cancellation in these cases the full matter content
from three generations of $27$-plets should survive to low energies.
the low energy matter content would then be extended to, \be
(\hat{Q}_i,\,\hat{u}^c_i,\,\hat{d}^c_i,\,\hat{L}_i,\,\hat{e}^c_i)
+(\hat{D}_i,\,\hat{\overline{D}}_i)+(\hat{S}_{i})+(\hat{H}^u_i)+(\hat{H}^d_i),\ee
where the $\hat{S}_{i}$, $\hat{H}^u_i$ and $\hat{H}^d_i$ are three
generations of doublets, whose quantum numbers will be like the
singlet, and up-, down-type Higgs fields and the $\hat{D}_i$ and
$\hat{\overline{D}}_i$ being $SU(3)_C$ triplets that reside in the
same $SU(5)$ multiplets as these Higgs-like states.

>>>>

@@

%%%--------------Inserted from introduction-------------------%%%%

Here we will refer to U(1) extensions of the MSSM, which solve the
$\mu$-problem as the USSM \cite{Cvetic:1995rj, Jain:1995cb,
  Nir:1995bu, Cvetic:1996mf, Cvetic:1997ky}.  The couplings for the
$U(1)^\prime$ gauge group should allow the following renormalisable
superpotential terms required in the USSM, \be W_{USSM} = y^U_{ij}
\hat{u}^c_i \hat{H}_u \hat{Q}_j + y^D_{ij} \hat{d}^c_i \hat{H}_d
\hat{Q}_j + y^E_{ij} \hat{e}^c_i \hat{H}_d \hat{L}_j + \lambda \hat{S}
\hat{H}_d \hat{H}_u, \ee with $i,j \in \{1,2,3\}$.

The $U(1)^\prime$ charges should allow for cancellations of gauge
anomalies.  The most elegant way to do this is to use an extra $U(1)$
gauge symmetry that can be obtained from the break down of the $E_6$
gauge symmetry which is anomaly free and have all matter that fill the
three generations of $27$-plet representations of $E_6$ survive down to
low energies.  Such models are often referred to in the literature as
$E_6$ inspired, and we will adopt this here.

@@

Here we will focus most on the special case where the gauge symmetry
is $U()_N$,under which the right handed neutrino remains
interaction-less. This is the case in the the exceptional
supersymmtric standard models (E$_6$SSM)
\cite{King:2005jy,King:2005my,Athron:2010zz}, and closely-related
variants\cite{Howl:2007zi, Braam:2009fi, Braam:2010sy, Hall:2011zq,
  Nevzorov:2012hs, Athron:2014pua}.  Since the right handed neutrino
has no gauge symmetry protecting it's mass from becoming extremely
heavy such models may explain the tiny observed masses of neutrinos
via the see-saw mechanism and the baryon asymmetry in the universe via
leptogenesis \cite{Hambye:2000bn,King:2008qb, King:2008gw}.  Recently
it has also been studied in the context of electroweak baryogenesis
\cite{Chao:2014hya}.

The gauge coupling running in the E$_6$SSM at the two-loop level leads
to unification more precisely than in the MSSM \cite{King:2007uj} or
in slightly modified scenarios two-step unification can take place
\cite{Howl:2007hq,Howl:2007zi}. If the exotic particles are light in
these models this can open up non-standard decays of the SM--like
Higgs boson \cite{Hall:2010ix,Nevzorov:2013tta,Athron:2014pua}. The
correct relic abundance could be obtained entirely through an almost
decoupled ``inert'' neutralino sector \cite{Hall:2009aj}.  However
this is no longer phenomenologically viable due to limits from direct
detection of dark matter
\cite{2011PhRvL.107m1302A,2012PhRvL.109r1301A, Akerib:2013tjd} and due
to a significant suppression of the decay of the lightest Higgs boson
into SM states, due to a new channel into inert singlinos opening up.

The impact of gauge kinetic mixing in the case where both of the extra
U(1)'s appearing from the breakdown of E$_6$ are present at low energy
was studied in \cite{Rizzo:2012rf}.  The E$_6$SSM was also included in
a studies looking at how first or second generation sfermion masses
can be used to constrain the GUT scale parameters \cite{Miller:2012vn}
and the renormalization of VEVs \cite{Sperling:2013eva,
  Sperling:2013xqa}.  The particle spectrum and collider signatures of
constrained version of the E$_6$SSM have been studied in a series of
papers,
\cite{Athron:2009ue,Athron:2009bs,Athron:2011wu,Athron:2012sq}. The
threshold corrections to the $\overline{DR}$ gauge and Yukawa
couplings in the E$_6$SSM were calculated and the numerical impact in
the constrained version examined \cite{Athron:2012pw}.
%%%%%%%%%%%%%%%%%%---inserted from introduction---%%%%%%%%%%%%%%%%5%%

With three generations of matter $27$-plet representations of $E_6$
surviving to low energies, the low energy matter content includes, \be
(\hat{Q}_i,\,\hat{u}^c_i,\,\hat{d}^c_i,\,\hat{L}_i,\,\hat{e}^c_i)
+(\hat{D}_i,\,\hat{\overline{D}}_i)+(\hat{S}_{i})+(\hat{H}^u_i)+(\hat{H}^d_i),\ee
where the $\hat{S}_{i}$, $\hat{H}^u_i$ and $\hat{H}^d_i$ are three
generations of doublets, whose quantum numbers will be like the
singlet, and up-, down-type Higgs fields and the $\hat{D}_i$ and
$\hat{\overline{D}}_i$ being $SU(3)_C$ triplets that reside in the
same $SU(5)$ multiplets as these Higgs-like states.
